Open-source tools in R for landscape ecology

### 2021-01-15 paper/paper.Rmd
:pencil: Add missing describtions to table and glossary

Glossary

\begin{tabularx}{\textwidth}{XX}

@@

CRAN & Comprehensive R Archive Network \\
CRS & Coordinate references systems \\
GIS & Geographical information system \\
NLM & Neutral landscape model \\
raster & Cell grid based data model \\
R packages & Software extension to R programming language \\
SDM & Species distribution modelling \\
vector & Points and lines based data model \\

@@

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

@@

Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
Since publicly available data is becoming more prominent, so are *R* packages to access them.
Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package *geodata* <!--(add reference) https://github.com/rspatial/geodata-->).
Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g. normalized difference vegetation index, disturbance intensity, topography). 
In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.

@@

\begin{tabularx}{450pt}{lrrr}

\hline
Task & R package & Describtion & Reference \\
\hline
\multirow{8}{*}{Spatial data} & raster & Raster data handling and analysis & \cite{Hijmans2019} \\
& terra & Basic data handling and analysis & \cite{R-terra} \\
& stars & Spatiotemporal raster data & \cite{Pebesma2019} \\
& fasterize & Polygon to raster conversion & \cite{Ross2020} \\
& rasterDT & Faster alternatives for some raster functions & \cite{OBrien2020} \\
& exactextractr & Summarize raster values over polygonal areas & \cite{Baston2020} \\
& sp & Vector data handling and analysis & \cite{Pebesma2005,Bivand2013} \\
& sf & Vector data handling and analysis & \cite{Pebesma2018} \\
\hline
\multirow{6}{*}{Spatial data download} & rnaturalearth & Download region and country data & \cite{South2017} \\
& elevatr & Download elevation data & \cite{R-elevatr} \\
& rgbif & Download biodiversity data & \cite{R-rgbif} \\
& BIEN & Download plant diversity, function and distribution data & \cite{Maitner2020} \\
& marmap & Download and manipulate bathymetric data & \cite{Pante2013} \\
& FedData & Download US federal data & \cite{R-FedData} \\
& getlandsat & Download satellite data from Landsat 8 & \cite{Chamberlain2018} \\
& MODIS & Download satellite data from MODIS & \cite{Mattiuzzi2020} \\
& sen2r & Download satellite data from Sentinel & \cite{R-sen2r} \\
\hline
\multirow{7}{*}{Creating maps} & mapview & Interactive viewing of spatial data & \cite{R-mapview} \\
& tmap & Create thematic maps & \cite{Tennekes2018} \\
& leaflet & Create interactive web maps & \cite{R-leaflet} \\
& cartography & Create cartographic maps & \cite{Giraud2016} \\
& ggspatial & Spatial extension for ggplot2 & \cite{Dunnington2020} \\
& rayshader & Create 2D and 3D data visualizations & \cite{Morgen-Wall2020} \\
\hline
\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019a} \\
& belg & Calculates the Boltzmann entropy & \cite{R-belgpaper} \\
& motif & Not on CRAN? & \cite{R-motif} \\
& geodiv & Quantify continuous landscape patterns & \cite{Smith2020} \\
\hline
\multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & Add describtion & \cite{Hijmans2017} \\
& sdm & Add describtion & \cite{Naimi2016} \\
& ecospat & Add describtion & \cite{Broennimann2020} \\
& biomod2 & Add describtion & \cite{Thuiller2020} \\
& PresenceAbsence & Add describtion & \cite{Freeman2008} \\
& zoon & Add describtion & \cite{Golding2018} \\
& adehabitatHR, adehabitatHS & Add describtion & \cite{Calenge2006} \\
& amt &  Add describtion &\cite{Signer2019} \\
\hline
\multirow{5}{*}{Connectivity} & lconnect & Add describtion & \cite{Mestre2019} \\
& Makurhini & Add describtion & \cite{Godinez-Gomez2020} \\
& grainscape & Add describtion & \cite{Chubaty2020} \\
& gdistance & Add describtion & \cite{vanEtten2017} \\
& samc & Add describtion & \cite{Marx2020} \\
\hline
\multirow{4}{*}{Landscape genetics} & raphs4lg & Add describtion & \cite{Savary2020} \\
& PopGenReport & Add describtion & \cite{Adamack2014,Gruber2015} \\
& HierDpart & Add describtion & \cite{Qin2019} \\
& GeNetIt & Add describtion & \cite{Murphy2010} \\
\hline
\multirow{4}{*}{various} & NLMR & Simulate neutral landscape models & \cite{Sciaini2018} \\
& RandomFields & Add describtion & \cite{Schlather2015} \\
& landscapetools & Utility functions for raster data & \cite{Sciaini2018}  \\
& grainchanger & Data aggregation methods for raster data & \cite{Graham2019} \\
\hline

>>>>

\begin{tabularx}{450pt}{lX}

@@

\multicolumn{2}{c}{Glossary} \\

API & Application programming interface \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive R Archive Network \\
CRS & Coordinate references systems \\
GIS & Geographical information system \\
landscapes & Mosaics of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
R packages & User-created software extension to R programming language \\
SDM & Species distribution modelling \\
vector & Points and lines based data model \\

@@

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

@@

Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
Since publicly available data is becoming more prominent, so are *R* packages to access them.
Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to downlad region and country data, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package *geodata* <!--(add reference) https://github.com/rspatial/geodata-->).
Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g. normalized difference vegetation index, disturbance intensity, topography). 
In landscape ecology, geostatistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape pattern; iii) interpolating data.

@@

\begin{tabularx}{450pt}{lrXr}
Task & R package & Describtion & Reference \\
\multirow{8}{*}{Spatial data} & raster & Raster data handling and analysis & \cite{Hijmans2019} \\
& terra & Raster (and some vector) data handling and analysis & \cite{R-terra} \\
& stars & Spatiotemporal raster data handling and analysis & \cite{Pebesma2019} \\
& fasterize & Polygon to raster conversion & \cite{Ross2020} \\
& rasterDT & Faster alternatives for some raster functions & \cite{OBrien2020} \\
& exactextractr & Summarize raster values over polygonal areas & \cite{Baston2020} \\
& sp & Vector data handling and analysis & \cite{Pebesma2005,Bivand2013} \\
& sf & Vector data handling and analysis & \cite{Pebesma2018} \\
\multirow{6}{*}{Spatial data download} & rnaturalearth & Download region and country data & \cite{South2017} \\
& elevatr & Download elevation data & \cite{R-elevatr} \\
& rgbif & Download biodiversity data & \cite{R-rgbif} \\
& BIEN & Download plant diversity, function and distribution data & \cite{Maitner2020} \\
& marmap & Download and manipulate bathymetric data & \cite{Pante2013} \\
& FedData & Download geospatial data from US federal sources & \cite{R-FedData} \\
& getlandsat & Download satellite data from Landsat 8 & \cite{Chamberlain2018} \\
& MODIS & Download satellite data from MODIS & \cite{Mattiuzzi2020} \\
& sen2r & Download satellite data from Sentinel & \cite{R-sen2r} \\
\multirow{7}{*}{Creating maps} & mapview & Interactive viewing of spatial data & \cite{R-mapview} \\
& tmap & Create thematic maps & \cite{Tennekes2018} \\
& leaflet & Create interactive web maps & \cite{R-leaflet} \\
& cartography & Create cartographic maps & \cite{Giraud2016} \\
& ggspatial & Spatial extension for ggplot2 & \cite{Dunnington2020} \\
& rayshader & Create 2D and 3D data visualizations & \cite{Morgen-Wall2020} \\
\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019a} \\
& belg & Calculate the Boltzmann entropy & \cite{R-belgpaper} \\
& motif & Pattern-based spatial analysis & \cite{R-motif} \\
& geodiv & Quantify continuous landscape patterns & \cite{Smith2020} \\
\multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & Methods for species distribution modeling & \cite{Hijmans2017} \\
& sdm & Species distribution models using individual and community-based approaches & \cite{Naimi2016} \\
& ecospat & Species distribution, niche quantification and community assembly & \cite{Broennimann2020} \\
& biomod2 & Species distribution modeling, ensemble of models and ensemble forecasting & \cite{Thuiller2020} \\
& PresenceAbsence & Evaluating of presence-absence models & \cite{Freeman2008} \\
& zoon & Reproducible and remixable species distribution modelling & \cite{Golding2018} \\
& adehabitatHR, adehabitatHS & Home range and habitat selection modelling & \cite{Calenge2006} \\
& amt & Manage and analyze animal movement data &\cite{Signer2019} \\
\multirow{5}{*}{Connectivity} & lconnect & Calculate landscape connectivity metrics & \cite{Mestre2019} \\
& Makurhini & Calculate fragmentation and landscape connectivity indices & \cite{Godinez-Gomez2020} \\
& grainscape & Calculate minimum planar graph and grains of connectivity models & \cite{Chubaty2020} \\
& gdistance & Distances and routes on geographical grids & \cite{vanEtten2017} \\
& samc & Functions for working with absorbing Markov chains & \cite{Marx2020} \\
\multirow{4}{*}{Landscape genetics} & graph4lg & Build graphs for landscape genetics analysis & \cite{Savary2020} \\
& PopGenReport & Framework to analyse population genetic data & \cite{Adamack2014,Gruber2015} \\
& HierDpart & Calculating and decomposing hierarchical diversity metrics & \cite{Qin2019} \\
& GeNetIt & Spatial graph-theoretic genetic gravity models & \cite{Murphy2010} \\
\multirow{4}{*}{various} & NLMR & Simulate neutral landscape models & \cite{Sciaini2018} \\
& RandomFields & Simulation and analysis of random fields & \cite{Schlather2015} \\
& landscapetools & Utility functions for raster data & \cite{Sciaini2018}  \\
& grainchanger & Data aggregation methods for raster data & \cite{Graham2019} \\

### 2021-01-15 paper/paper.tex
:pencil: Add missing describtions to table and glossary

One successful example of an open-source project is the \emph{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.

@@

Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g.~normalized difference vegetation index, disturbance intensity, topography).
In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.

>>>>

One successful example of an open-source project is the \emph{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.

@@

Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to downlad region and country data, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g.~normalized difference vegetation index, disturbance intensity, topography).
In landscape ecology, geostatistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape pattern; iii) interpolating data.
